Valuing Real Estate in Seattle Using Property Characteristics: A Machine Learning Approach

### 2024-11-28 paper/paper.qmd
Updating LLM usage and paper

library(tidyverse)
library(rsample)
library(kableExtra)
library(arrow)
library(stargazer)
library(ggpubr)
library(modelsummary)
library(dagitty)
library(ggdag)
```

@@

$$
\text{price} = \beta_0 + \beta_1 \cdot \text{beds} + \beta_2 \cdot \text{sqft} + \beta_3 \cdot \text{hoa\_month} + \beta_4 \cdot (\text{sqft} \cdot \text{beds})
$$

@@

The effect of adding one more bedroom ($\text{beds}$) on `price`, holding other factors constant (except for the interaction term).

**$\beta_2$: Coefficient for `sqft`**

The effect of increasing square footage ($\text{sqft}$) by one unit on `price`, holding other factors constant (except for the interaction term).

**$\beta_3$: Coefficient for `hoa_month`**

The effect of a one-unit increase in monthly homeowner's association fees (`hoa_month`) on `price`, holding other factors constant.

**$\beta_4$: Interaction between `sqft` and `beds`**

Represents how the relationship between $\text{sqft}$ and `price` changes depending on the number of bedrooms ($\text{beds}$).

@@

@@

@tbl-summ shows the values of the model's coefficients mentioned in the setup. The intercept of -341,047.42 suggests that, without any additional predictors, property prices would begin at this negative baseline, though this value has limited standalone interpretation. The coefficient for beds is -159,574.45, meaning that, holding other factors constant, each additional bedroom is associated with a significant reduction in price. However, this effect is modified by the interaction term beds $\times$ sqft, which has a negative coefficient of -32.64, indicating that larger properties with more bedrooms have diminishing price returns. The sqft variable has a strong positive impact (1152.06 per additional square foot), highlighting size as a key driver of value. The hoa_month coefficient (267.83) shows a modest positive association with price, possibly reflecting higher costs in premium communities. With an adjusted $R^2$ of 0.6543, the model explains 65.3\% of the variance in property prices, and an RMSE of 719,452.71, the summary suggests a decent fit.

@@

```{r #tbl-summ}
modelsummary(list(
  "Final Model" = final_model), output = "kableExtra",
  fmt = "%.2f",
  gof_omit = "BIC|AIC") |>
  column_spec(2, width = "2.5in")
```

```{r fig-dag, fig.align='center', fig.width=4, fig.height=3, fig.cap='Multiple Causal Relationships Among Model Variables'}
dagify(
  price ~ beds,
  price ~ sqft,
  price ~ baths,
  price ~ hoa,
  sqft ~ beds,
  sqft ~ baths,
  baths ~ beds,
  hoa ~ sqft,
  coords = list(
    x = c(beds = 0.5, sqft = 1, baths = 0, price = 2, hoa = 1.5),
    y = c(beds = 0, sqft = 1, baths = 1, price = 0, hoa = 1.5)
  )
) |>
  ggdag() +
  theme_void()
```

@@

The linear model was created using base R [@R] to fit the data and generate predictions, while the MLMetrics package [@mlmetrics] was used to evaluate performance. A train-test split of 70:30 was created with the rsample package [@rsample], with the model trained on the larger portion of the data and validated using out-of-sample testing on the remaining data. Key evaluation metrics included the Root Mean Square Error (RMSE) and the $R^2$ score. The model achieved an RMSE of 1,850,720, indicating the average prediction error in dollar terms, and an $R^2$ score of approximately 0.281, reflecting the proportion of variance in property prices explained by the model. Detailed diagnostics and further information are provided in Appendix B.

@@

Several alternative models were considered before selecting the final one. The first, a “full model,” included all variables in the cleaned dataset, achieving an RMSE of 1,864,966 and an $R^2$ score of 0.269. While this model used all available information, many variables were statistically insignificant. A reduced model was then tested by retaining only significant variables, resulting in an RMSE of 1,865,188 and an $R^2$ score of 0.269. Though concise, it performed similarly to the full model. The final model, incorporating an interaction term, outperformed both alternatives with an RMSE of 1,850,720 and an $R^2$ score of 0.281.

@@

```{r fig-scatter, fig.cap='Large Uncertainty in the Values of Luxury Homes and Mansions'}
testing_data |>
  ggplot(aes(x = predicted_price / 10^6, y = price / 10^6)) +
  geom_point(color = "#519C7C", size = 1) +
  geom_function(fun = function(x){x}, linetype = "dashed") +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  scale_y_continuous(breaks = seq(0, 20, 2)) +
  labs(x = "Predicted Price ($ millions)", y = "Actual Price ($ millions)")
```

@@

```{r fig-col, fig.cap="Highest Average Predicted Listing Price in Seattle's 85 Acre Gated Community"}
testing_data |>
  select(neighbourhood, predicted_price) |>
  group_by(neighbourhood) |>
  summarise(mean = mean(predicted_price / 10^6)) |>
  arrange(desc(mean)) |>
  head(n = 5) |>
  ggplot(aes(x = reorder(neighbourhood, mean), y = mean)) +
  geom_col(position = "dodge", fill = "#519C7C", color = "black") +
  scale_y_continuous(breaks = seq(0, 5, 0.5)) +
  coord_flip() +
  labs(x = "Neighbourhood", y = "Mean Predicted Price ($ millions)")
```

@@

```{r fig-box, fig.cap='40 Year Old Homes Among the Lowest in Actual and Predicted Values'}
testing_data |>
  group_by(property_age) |>
  summarise(mean_price = mean(price) / 10^6,
            mean_predicted = mean(predicted_price) / 10^6) |>
  pivot_longer(names_to = "price_type", cols = c(mean_price, mean_predicted),
               values_to = "value") |>
  mutate(price_type = recode(price_type, 
                             "mean_price" = "Actual", 
                             "mean_predicted" = "Predicted")) |>
  ggplot(aes(x = property_age, y = value, color = price_type)) +
  geom_point(size = 0.75) +
  scale_color_manual(values = c("black", "#519C7C")) +
  geom_smooth(method = "lm", formula = y ~ I(x^2) + x, se = FALSE) +
  facet_wrap(vars(price_type)) +
  theme(legend.position = "top") +
  labs(x = "Property Age", y = "Average Price ($ millions)", color = "Price Type")
```

>>>>

library(tidyverse)
library(rsample)
library(kableExtra)
library(arrow)
library(stargazer)
library(ggpubr)
library(modelsummary)
library(dagitty)
library(ggdag)
library(sf)
```

@@

$$
\text{price} = \beta_0 + \beta_1 \cdot \text{beds} + \beta_2 \cdot \text{baths} + \beta_3 \cdot \text{sqft} + \beta_4 \cdot \text{hoa\_month} + \beta_5 \cdot (\text{sqft} \cdot \text{beds})
$$

@@

The effect of adding one more bedroom on price, holding other factors constant (except for the interaction term).

**$\beta_2$: Coefficient for `baths`**

The effect of adding one more bathroom on price, holding other factors constant.

**$\beta_3$: Coefficient for `sqft`**

The effect of increasing square footage by one unit on price, holding other factors constant (except for the interaction term).

**$\beta_4$: Coefficient for `hoa_month`**

The effect of a one-unit increase in monthly homeowner's association fees on price, holding other factors constant.

**$\beta_5$: Interaction between `sqft` and `beds`**

Represents how the relationship between $\text{sqft}$ and `price` changes depending on the number of bedrooms.

@@

```{r fig-dag, fig.width=4, fig.height=2.8, fig.cap='Multiple Causal Relationships Among Model Variables'}
dagify(
  price ~ beds,
  price ~ sqft,
  price ~ baths,
  price ~ hoa,
  sqft ~ beds,
  sqft ~ baths,
  baths ~ beds,
  hoa ~ sqft,
  coords = list(
    x = c(beds = 0.5, sqft = 0.9, baths = 0, price = 2, hoa = 1.6),
    y = c(beds = 0, sqft = 0.9, baths = 1, price = 0, hoa = 1.2)
  )
) |>
  ggdag() +
  theme_void()
```

@@

@tbl-summ shows the values of the model's coefficients mentioned in the setup. The intercept of −66,054.23 suggests that, without any additional predictors, property prices would begin at this value, though it has limited standalone interpretation. The coefficient for `beds` is 55,091.77, meaning that, holding other factors constant, each additional bedroom is associated with an addition in price. However, this effect is modified by the interaction term `beds` $\times$ `sqft`, which has a negative coefficient of −38.66, indicating that larger properties have diminishing prices. The coefficient for `baths` is 41,740.69, meaning that for each additional bathroom, the predicted value of the price increases by around \$41,740.69, holding all other variables constant. This suggests that bathrooms have a strong positive impact on the value of the property. The `sqft` variable has a strong positive impact (592.23 per additional square foot), highlighting size as a key driver of value. The `hoa_month` coefficient (66.39) shows a modest positive association with price, possibly reflecting higher costs in premium communities. With an adjusted $R^2$ of 0.716 and an RMSE of 331,563.51, the model has a decent fit.

@@

```{r #tbl-summ}
modelsummary(list(
  "Final Model" = final_model), output = "kableExtra",
  fmt = "%.2f", gof_map = c("nobs", "r.squared", "adj.r.squared", "rmse")) |>
  column_spec(2, width = "2.5in")
```

@@

The linear model was created using R [@R] to fit the data and generate predictions, while the MLMetrics package [@mlmetrics] was used to evaluate performance. A train-test split was created with the rsample package [@rsample], with the model trained on the 70\% of the data and validated using out-of-sample testing on the remaining 30\%. Key evaluation metrics included the Root Mean Square Error (RMSE) and $R^2$ The model achieved an RMSE of 331,564, indicating the average prediction error in dollar terms, and $R^2$ of approximately 0.716, reflecting the proportion of variance in property prices explained by the model. Further diagnostics and information are provided in Appendix B.

@@

Several alternative models were considered before selecting the final one. The first, a “full model,” included all variables in the cleaned dataset, achieving an RMSE of 1,864,966 and an $R^2$ score of 0.269. While this model used all available information, many variables were statistically insignificant. A reduced model was then tested by retaining only significant variables, resulting in an RMSE of 1,865,188 and an $R^2$ score of 0.269. Though concise, it performed similarly to the full model. The final model, incorporating an interaction term, outperformed both alternatives with an RMSE of 1,850,720 and an $R^2$ of 0.281.

@@

```{r fig-scatter, fig.cap='Large Uncertainty in the Predicted Values of Luxury Homes'}
testing_data |>
  ggplot(aes(x = predicted_price / 10^6, y = price / 10^6)) +
  geom_point(color = "#519C7C", size = 1, shape = 15) +
  geom_abline(linetype = "dashed") +
  scale_x_continuous(breaks = seq(0, 20, 1)) +
  scale_y_continuous(breaks = seq(0, 20, 2)) +
  labs(
    x = "Predicted Price ($ millions)",
    y = "Actual Price ($ millions)"
  )
```

@@

```{r fig-box, fig.cap="Single Family Homes Valued the Highest Among All Property Types"}
testing_data |>
  pivot_longer(
    cols = c(price, predicted_price), 
    names_to = "price_type", 
    values_to = "value"
  ) |>
  mutate(price_type = recode(price_type, "price" = "Actual", 
                             "predicted_price" = "Predicted")) |>
  ggplot(aes(x = property_type, y = value / 10^6, fill = price_type)) +
  geom_boxplot(
    position = position_dodge(0.6), 
    color = "black", 
    width = 0.5,
    outlier.shape = 15, 
    outlier.size = 0.5
  ) +
  scale_y_continuous(limits = c(0, 7)) +
  coord_flip() +
  scale_fill_manual(values = c("grey70", "#519C7C")) +
  theme(legend.position = "top") +
  labs(
    x = "Property Type", 
    y = "Price ($ millions)", 
    fill = "Price Type"
  )
```

@@

```{r fig-age, fig.cap='40 Year Old Homes Among the Lowest in Actual and Predicted Values'}
testing_data |>
  group_by(property_age) |>
  summarise(
    mean_price = mean(price) / 10^6,
    mean_predicted = mean(predicted_price) / 10^6
  ) |>
  pivot_longer(
    names_to = "price_type", 
    cols = c(mean_price, mean_predicted),
    values_to = "value"
  ) |>
  mutate(price_type = recode(price_type, 
                             "mean_price" = "Actual", 
                             "mean_predicted" = "Predicted")) |>
  ggplot(aes(x = property_age, y = value, color = price_type)) +
  geom_point(size = 0.75, shape = 15) +
  scale_color_manual(values = c("black", "#519C7C")) +
  geom_smooth(method = "lm", formula = y ~ I(x^2) + x, se = FALSE) +
  facet_wrap(vars(price_type)) +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Property Age",
    y = "Average Price ($ millions)",
    color = "Price Type"
  )
```

## Result 4

```{r fig-map, fig.cap="Highest Average Predicted Listing Price in Seattle's 85 Acre Gated Community"}
zip_codes <- st_read("../data/spatial_data/zip-codes.geojson", quiet = TRUE) |>
  rename(zip_code = ZCTA5CE10) |> 
  select(zip_code, geometry)

testing_data |>
  group_by(zip_code) |>
  summarise(
    mean_price = mean(price) / 10^6,
    mean_predicted = mean(predicted_price) / 10^6,
    .groups = "drop"
  ) |>
  pivot_longer(
    names_to = "price_type", 
    cols = c(mean_price, mean_predicted),
    values_to = "value"
  ) |>
  mutate(price_type = recode(price_type, 
                             "mean_price" = "Actual", 
                             "mean_predicted" = "Predicted")) |>
  merge(zip_codes, by = "zip_code") |>
  st_sf() |>  # Convert to sf object
  ggplot() +
  geom_sf(aes(fill = value)) +
  coord_sf(
    xlim = c(-122.5369, -122.1166), 
    ylim = c(47.44308, 47.79792),
    expand = FALSE
  ) +
  scale_fill_distiller(palette = "Greens", direction = 1) +
  facet_wrap(vars(price_type)) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(fill = "Price \n($ millions)")
```
